---
title: "Chapter 5"
author: "Christine Iyer"
date: "March 19, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#1. Read in data and convert to time series. 

```{r}
library(forecast)
library(zoo)
library(knitr)
#read in data
ridership <- read.csv("Amtrak.csv")
##convert data to a time series
ridership.ts <- ts(ridership$Ridership_in_thousands, start = c(1991,1), end = c(2004, 3), frequency = 12)
tail(ridership)
```

#2. Visualize for trend and seasonality

###Trend
```{r}
#trend line
ridership.lm <- tslm(ridership.ts ~ trend + I(trend^2)) 
par(mfrow = c(2, 1)) 
#simple plot
plot(ridership.ts, xlab = "Time", ylab = "Ridership", ylim = c(1300, 2300), bty = "l")
#plot fitted trend line
lines(ridership.lm$fitted, lwd = 2) 
#zoom into 4 years
#first create window
ridership.ts.zoom <- window(ridership.ts, start = c(1997, 1), end = c(2000, 12))
#plot zooned in data
plot(ridership.ts.zoom, xlab = "Time", ylab = "Ridership", ylim = c(1300, 2300), bty = "l") 
```

**Using centered moving average to visualize trend, along with linear trend line**

```{r}
# moving average by one year
ma.centered <- ma(ridership.ts, order = 12) 
```


```{r}
plot(ridership.ts, xlab = "Time", ylab = "Ridership", ylim = c(1300, 2300), bty = "l")
#plot fitted trend line
lines(ridership.lm$fitted, lwd = 2, col = 4) 
lines(ma.centered, lwd = 2, col = 3) 
legend(1994,2300, c("Ridership","Trend", "Centered Moving Average"), lty=c(1,1,1), lwd=c(2,2,2), col = c(1,4,3), bty = "o")

```


###Seasonality

**Look for seasonality**

```{r}
library(ggplot2)
ggseasonplot(ridership.ts, ylab = "Amtrak Ridership", main = "Seasonal Plot for Amtrak Ridership", lwd = 2) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank())
```

**Ridership boken out by month**

```{r}
par(oma = c(0, 0, 0, 2))
xrange <- c(1992,2004)
yrange <- range(ridership.ts)
plot(xrange, yrange, type="n", xlab="Year", ylab="Amtrak Ridership", bty="l", las=1)
colors <- rainbow(12) 
linetype <- c(1:12) 
plotchar <- c(1:12)
axis(1, at=seq(1992,2004,1), labels=format(seq(1992,2004,1)))
for (i in 1:12) { 
  currentMonth <- subset(ridership.ts, cycle(ridership.ts)==i)
  lines(seq(1992, 1992 +length(currentMonth)-1,1), currentMonth, type="b", lwd=1,
      lty=linetype[i], col=colors[i], pch=plotchar[i]) 
} 
title("Ridership Broken Out by Month")
legend(2002.35, 80, 1:12, cex=0.8, col=colors, pch=plotchar, lty=linetype, title="Month", xpd=NA)

```

**There appears to be some seasonality**

###3. Forecast with moving average after double differencing 

**Double differencing**

```{r}
diff.twice.ts <- diff(diff(ridership.ts, lag = 12), lag = 1)
```

**Trailing moving average of double differenced Amtrak data**

```{r}
#create validation period length
nValid <- 36
#create training period length
nTrain <- length(diff.twice.ts) - nValid 
#create training period window
train.ts <- window(diff.twice.ts, start = c(1992, 2), end = c(1992, nTrain + 1)) 
#create validation period window
valid.ts <- window(diff.twice.ts, start = c(1992, nTrain + 2), end = c(1992, nTrain + 1 + nValid))
#trailing moving average of training period
ma.trailing <- rollmean(train.ts, k = 12, align = "right")
forecast.ma.trailing <- forecast(ma.trailing, h = nValid, level = 0)
accuracy(forecast.ma.trailing, valid.ts)
```

**4. Simple Exponential Smoothing using only seasonal differencing**

```{r}
diff.ts <- diff(ridership.ts, lag = 12)
#create validation period length
nValid <- 36
#create training period length
nTrain <- length(diff.ts) - nValid 
#create training period window
train.ts <- window(diff.ts, start = c(1992, 2), end = c(1992, nTrain + 1)) 
#create validation period window
valid.ts <- window(diff.ts, start = c(1992, nTrain + 2), end = c(1992, nTrain + 1 + nValid))
```

ANN, $\alpha = 0.2$
```{r}
#simple exponential smoothing with no trend and no seasonality model, alpha = 0.2
ses1 <- ets(train.ts, model = "ANN", alpha = 0.2)
#forecast the above ses model. The forecast is flat because the trend and seasonality are not configured in. 
ses.pred <- forecast(ses1, h = nValid, level = 0)
#accuracy of ses, ANN forecast with alpha = .2
accuracy(ses.pred, valid.ts)
ses1
```

ANN, $\alpha = null$
```{r}
#simple exponential smoothing with no trend and no seasonality model, ANN, no predetermined alpha
ses2 <- ets(train.ts, model = "ANN")
#forecast the above ses model. The forecast is flat because the trend and seasonality are not configured in. 
ses.pred2 <- forecast(ses2, h = nValid, level = 0)
#accuracy of ses, ANN forecast with alpha = null
accuracy(ses.pred2, valid.ts)
ses2
```


**When no model determined, the function chooses ANN**

ZZZ, $\alpha = 0.2$
```{r}
#simple exponential smoothing with no trend and no predetermined model, alpha = 0.2
ses3 <- ets(train.ts, model = "ZZZ", alpha = 0.2)
#forecast the above ses model. The forecast is flat because the trend and seasonality are not configured in. 
ses.pred3 <- forecast(ses3, h = nValid, level = 0)
#accuracy of ses, ANN forecast with alpha = .2
accuracy(ses.pred3, valid.ts)
ses3
```

**Again, model chooses ANN and alpha of 0.468**
ZZZ, $\alpha = null$

```{r}
#simple exponential smoothing with no trend and no seasonality model, alpha  is blank
ses4 <- ets(train.ts, model = "ZZZ")
#forecast the above ses model. The forecast is flat because the trend and seasonality are not configured in. 
ses.pred4 <- forecast(ses4, h = nValid, level = 0)
#accuracy of ses, ANN forecast with alpha = blank
accuracy(ses.pred4, valid.ts)
ses4

```


###Advanced exponential smoothing using Holt Winters.

**Since this model is apppropriatefor series with trend and seasonality, no differencing is needed.** 

```{r}
#create validation period length
nValid <- 36
#create training period length
nTrain <- length(ridership.ts) - nValid 
#create training period window
train.ts <- window(ridership.ts, start = c(1993, 1), end = c(1992, nTrain + 1)) 
#create validation period window
valid.ts <- window(ridership.ts, start = c(199, nTrain + 2), end = c(1993, nTrain + 1 + nValid))
hwRidership <- HoltWinters(train.ts)
forecasthwRIdership <- forecast(hwRidership, h = nValid, level = 0)
accuracy(forecasthwRIdership, valid.ts)
hwRidership
```
