---
title: "CopyFormulas"
author: "Christine Iyer"
date: "February 17, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(forecast)
```

###Load the Data

```{r}
appliance <- read.csv("ApplianceShipments.csv", stringsAsFactors = FALSE) 
head(appliance)
tail(appliance)
```


###Data is out of order
**remove the dash between year and quarter**

```{r}
# Split the column Quarter based on the "-"
col1Split <- strsplit(appliance$Quarter,"-")
head(col1Split)
```

### Create a data.frame with three columns:

```{r}
tempDF <- data.frame(matrix(unlist(col1Split), nrow=length(col1Split), byrow=T), appliance$Shipments) 
tempDF
```

###Order according to year, then quarter

```{r}
# Order tempDF 
tempDF <- tempDF[order(tempDF$X2,tempDF$X1),]
# See if the sorting worked
tempDF
```

###Put the quarter and year back together

```{r}
m <- paste0(tempDF$X1,"-",tempDF$X2)
m
```

###Create new DF with just the 2 columns

```{r}
newDF <- data.frame(m, tempDF$appliance.Shipments)
newDF
```

# Rename the first and second column

```{r}
names(newDF)[1] <- "Quarter"
names(newDF)[2] <- "Shipments"
head(newDF)
```

###Convert to time series

```{r}
appliance.ts <- ts(newDF$Shipments, start = c(1985,1), end = c(1989, 4), freq = 4)
```

###Plot

```{r}
plot(appliance.ts, xlab = "Year", ylab = "Shipments (millions of US$", ylim = c(3000, 5000), bty = "l")
```

###Trend

```{r}
appliance.lm <- tslm(appliance.ts ~ trend + I(trend^2))
par(mfrow = c(2, 1))
plot(appliance.ts, xlab = "Year", ylab = "Shipments", ylim = c(3000, 5000), bty = "l") 
lines(appliance.lm$fitted, lwd = 2)
```

###Zoom helps suppress seasonality

```{r}
appliance.ts.zoom <- window(appliance.ts, start = c(1986,1), end = c(1988, 4))
plot(appliance.ts.zoom, xlab = "Time", ylab = "Shipments (millions of US$", ylim = c(3000, 5000), bty = "l")
```

###Partition the data into training and validation periods for forecasting 
**One year validation**

```{r}
applianceValidation <- 4
applianceTraining <- length(appliance.ts) - applianceValidation
```

##Put partitions in windows

###Training Window

```{r}
applianceTrainWindow <- window(appliance.ts, start = c(1985, 1), end = c(1985, applianceTraining))
```

###Validation Window

```{r}
applianceValidationWindow <- window(appliance.ts, start = c(1985, applianceTraining + 1), end = c(1985, applianceValidation + applianceTraining))
appliance.lm <- tslm(applianceTrainWindow ~ trend + I(trend^2))
appliance.lm.pred <- forecast(appliance.lm, h = applianceValidation, level = 0)
plot(appliance.lm.pred, ylim = c(3000, 5000), xlab = "Year", ylab = "Shipments",bty = "l", xaxt = "n", xlim = c(1985,1989.25), main ="", flty = 2)
labels = format(seq(1985, 1989,1))
axis(1, at = seq(1985, 1989, 1), labels = format(seq(1985, 1989, 1)))
lines(appliance.lm$fittled, lwd = 2)
lines(applianceValidationWindow)
```

```{r}

```

```{r}

```

```{r}

```


 lines(ridership.lm$fitted, lwd = 2) 
lines(valid.ts)



