---
title: "Predictive Analytics Assignment 2"
author: "Christine Iyer"
date: "February 8, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Souvenir Sales: The file SouvenirSales.xls contains monthly sales for a souvenir shop at a beach resort town in Queensland, Australia, between 1995 and 2001. Back in 2001, the store wanted to use the data to forecast sales for the next 12 months (year 2002). They hired an analyst to generate forecasts. The analyst first partitioned the data into training and validation periods, **with the validation period containing the last 12 months of data (year 2001)**. She then fit a forecasting model to sales, using the training period. Partition the data into the training and validation periods as explained above. 

```{r}
library(knitr)
library(forecast)
```

```{r}
souvenir <- read.csv("SouvenirSales.csv", header = TRUE, stringsAsFactors = FALSE)
kable(head(souvenir))
souvenir.ts <- ts(souvenir$Sales, start = c(1995, 1), frequency = 12)
summary(souvenir.ts)
```

```{r}
#1 validation
nValid <- 12 
#2 training
nTrain <- length(souvenir.ts) - nValid 
#3 training window
train.ts <- window(souvenir.ts, start = c(1995, 1), end = c(1995, nTrain)) 
# validation window
valid.ts <- window(souvenir.ts, start = c(1995, nTrain + 1), end = c(1995, nTrain + nValid)) 
#5 trend line
souvenir.lm <- tslm(train.ts ~ trend + I(trend^2)) 
#6 forecast on training prediction
souvenir.lm.pred <- forecast(souvenir.lm, h = nValid, level = 0)
#7 plot forecast
plot(souvenir.lm.pred, ylim = c(1660, 105000), ylab = "souvenir", xlab = "Time", bty = "l", xaxt = "n", xlim = c(1995,2002.25), main ="", flty = 2)
#8 axis
axis(1, at = seq(1995, 2002, 1), labels = format(seq(1995, 2002, 1)))
#9 trend line
lines(souvenir.lm$fitted, lwd = 2)
#10 validation line
lines(valid.ts)
```

**code for computing predictive measures**

```{r}
accuracy(souvenir.lm.pred$mean, valid.ts)
```

**code for calculating the forecast errors in the validation period, i.e., the residuals**

```{r}
#what's included in the forecast
names(souvenir.lm.pred) 
#the model's residuals are the forecast errors in the training period
souvenir.lm.pred$residuals 
#subtracting the model's mean (or forecasts) from valid.ts (or actuals) in the validation period gives the forecast errors in the validation period
valid.ts - souvenir.lm.pred$mean
```


**code for creating the histogram of the forecast errors in each of 7 equally sized bins**
```{r}
hist(souvenir.lm.pred$residuals, ylab = "Frequency", xlab = "Forecast Error", bty = "l", main ="") 
```

**forecast with 95% confidence interval**

```{r}
#6 forecast on training prediction
souvenir.lm.pred <- forecast(souvenir.lm, h = nValid, level = 95)
#7 plot forecast
plot(souvenir.lm.pred, ylim = c(1660, 105000), ylab = "souvenir", xlab = "Time", bty = "l", xaxt = "n", xlim = c(1995,2002.25), main ="", flty = 2)
#8 axis
axis(1, at = seq(1995, 2002, 1), labels = format(seq(1995, 2002, 1)))
#9 trend line
lines(souvenir.lm$fitted, lwd = 2)
#10 validation line
lines(valid.ts)
```

To create a normal probability plot in R, put the forecast error into the function qqnorm.

```{r}
qqnorm(souvenir.lm.pred$residuals)
quantile(souvenir.lm.pred$residuals)
```

```{r}
souvenir.ets.AAN <- ets(souvenir.ts, model = "AAN") #Fit Model 1 to the time series. 
souvenir.ets.MMN <- ets(souvenir.ts, model = "MMN", damped = FALSE) #Fit Model 2. 
souvenir.ets.MMdN <- ets(souvenir.ts, model = "MMN", damped = TRUE) #Fit Model 3. 
souvenir.ets.AAN.pred <- forecast(souvenir.ets.AAN, h = 15, level = c(0.2, 0.4, 0.6, 0.8)) 
souvenir.ets.MMN.pred <- forecast(souvenir.ets.MMN, h = 15, level = c(0.2, 0.4, 0.6, 0.8)) 
souvenir.ets.MMdN.pred <- forecast(souvenir.ets.MMdN, h = 15, level = c(0.2, 0.4, 0.6, 0.8)) 
par(mfrow = c(1, 3)) #This command sets the plot window to show 1 row of 3 plots. 
plot(souvenir.ets.AAN.pred, xlab = "Month", ylab = "souvenir (in millions)", ylim = c(1660, 105000)) 
plot(souvenir.ets.MMN.pred, xlab = "Month", ylab="souvenir (in millions)", ylim = c(1660, 105000)) 
plot(souvenir.ets.MMdN.pred, xlab = "Month", ylab="souvenir (in millions)", ylim = c(1660, 105000))
```

     

**code for computing the predictive measures for rollforward one-month-ahead forecasts**

```{r}
fixed.nValid <- 12
fixed.nTrain <- length(souvenir.ts) - fixed.nValid
stepsAhead <- 1
error <- rep(0, fixed.nValid - stepsAhead + 1)
percent.error <- rep(0, fixed.nValid - stepsAhead + 1)

for(j in fixed.nTrain:(fixed.nTrain + fixed.nValid - stepsAhead)) { 
  train.ts <- window(souvenir.ts, start = c(1995, 1), end = c(1995, j))
  valid.ts <- window(souvenir.ts, start = c(1995, j + stepsAhead), end = c(1995, j + stepsAhead)) 
naive.pred <- naive(train.ts, h = stepsAhead)   
error[j - fixed.nTrain + 1] <- valid.ts - naive.pred$mean[stepsAhead]
percent.error[j - fixed.nTrain + 1] <- error[j - fixed.nTrain + 1]/ valid.ts 
}


```


**Mean Absolute Error**
```{r, echo=FALSE}
kable( mean(abs(error)))
```


**Mean Absolute Squared Error**
```{r, echo=FALSE}
kable(sqrt(mean(error^2)))
```


**Mean Absolute Percent Error**
```{r, echo=FALSE}
kable(mean(abs(percent.error)))
```



**code for computing naive and seasonal naive forecasts and their predictive measures**

```{r}
fixed.nValid <- 12 
fixed.nTrain <- length(souvenir.ts) - fixed.nValid
train.ts <- window(souvenir.ts, start = c(1995, 1), end = c(1995, fixed.nTrain)) 
valid.ts <- window(souvenir.ts, start = c(1995, fixed.nTrain + 1), end = c(1995, fixed.nTrain + fixed.nValid)) 
```

```{r}
naive.pred <- naive(train.ts, h = fixed.nValid)
snaive.pred <- snaive(train.ts, h = fixed.nValid)
```

**Accuracy of Naive Forecast**

```{r}
kable(accuracy(naive.pred, valid.ts)) 
```

**Accuracy of Seasonal Naive Forecast**

```{r}
kable(accuracy(snaive.pred, valid.ts))
```


(a) Why was the data partitioned? 

(b) Why did the analyst choose a 12-month validation period? 

(c) What is the naive forecast for the validation period? (assume that you must provide forecasts for 12 months ahead) 

(d) Compute the RMSE and MAPE for the naive forecasts. 

(e) Plot a histogram of the forecast errors that result from the naive forecasts (for the validation period). Plot also a time plot for the naive forecasts and the actual sales numbers in the validation period. What can you say about the behavior of the naive forecasts? 

(f) The analyst found a forecasting model that gives satisfactory performance on the validation set. What must she do to use the forecasting model for generating forecasts for year 2002? 

2. Forecasting Shampoo Sales: The file ShampooSales.xls contains data on the monthly sales of a certain shampoo over a threeyear period.

If the goal is forecasting sales in future months, which of the following steps should be taken? (choose one or more) 

*partition the data into training and validation periods 

*examine time plots of the series and of model forecasts only for the training period

*look at MAPE and RMSE values for the training period 

*look at MAPE and RMSE values for the validation period 

*compute naive forecasts

