
---
title: "Smoothing Methods"
author: "Christine Iyer"
date: "March 9, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

```{r}
library(forecast)
library(zoo)
library(knitr)
#read in data
ridership <- read.csv("Amtrak.csv")
##convert datato a time series
ridership.ts <- ts(ridership$Ridership_in_thousands, start = c(1991,1), end = c(2004, 3), frequency = 12)
tail(ridership)
```

###2 Types of Moving Averages
**The difference between the two, is the placement of the averaging window over the time series.** 

1. Centered Moving Average

*Averaging operation can suppress seasonality and noise, making the trend more visible.

*If window width is 5, the moving average at t = 3 means averaging time points 1,2,3,4,5. At time point t = 4, the moving average is the average of 2,3,4,5,6. 

*Choosing the window length should be the length of the seasonal cycle. W = 12 for Amtrak data.

*Computed by averaging across data in the past and future of a given time point, therefore not useful for forecasting. 

*If trend and seasonality, don't use this method, unless just want to see the global trend. 

**Centered moving average uses ma() in forecast package.**

```{r}
# moving average by one year
ma.centered <- ma(ridership.ts, order = 12) 
```


2. Trailing Moving Average

*Useful for forecasting

*Window of width W is placed over the most recent W values of the time series. The k-step ahead forecast is then the average of these W values. 

*The lag makes this method not work for seasonal and trend data. According to textbook, "Seasons with high ridership are under-forecasted, and seasons with low ridership are over-forecasted. A similar issue arises when forecasting a series with a trend: the moving average "lags behind", thereby under-forecasting in the presence of an increasing trend and over-forecasting in the presence of a decreasing trend." But de-trending and de-seasonalizing can be done with regression models, advanced exponential smoothing methods, and differencing. After detrended, etc, forecast, then add back. 

**To do trailing moving aveage, use rollmean() in zoo package**

```{r}
#trailing moving average of entire time series
ma.trailing <- rollmean(ridership.ts, k = 12, align = "right")

```

**Plotting the time series and moving averages**
[graphical aids](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/legend.html)
```{r}
#plot the whole time series
plot(ridership.ts, ylim = c(1300, 2200), ylab = "Ridership", xlab = "Time", bty = "l", xaxt = "n", xlim = c(1991,2004.25), main ="")
#label x axis
axis(1, at = seq(1991, 2004.25, 1), labels = format(seq(1991, 2004.25, 1))) 
#add moving average line, color 3 = green, lwd = straight line
lines(ma.centered, lwd = 2, col = 3) 
#add trailing moving average line, color 4 = blue, lty = 2, dashed line
#lwd = line thickness
lines(ma.trailing, lwd = 2, lty = 2, col = 4) 
#legend for all three lines
#bty = "o", puts a box around the legend, and "n", no box around it.
legend(1994,2200, c("Ridership","Centered Moving Average", "Trailing Moving Average"), lty=c(1,1,2), lwd=c(1,2,2), col = c(1, 3:4), bty = "o")

```

**Trailing Moving Average Forecaster**

```{r}
#3 year validation period window
nValid <- 36
#training period
nTrain <- length(ridership.ts) - nValid
#training window
train.ts <- window(ridership.ts, start = c(1991, 1), end = c(1991, nTrain))
#validation window
valid.ts <- window(ridership.ts, start = c(1991, nTrain + 1), end = c(1991, nTrain + nValid))
#trailing moving average of training period
ma.trailing <- rollmean(train.ts, k = 12, align = "right")
#last value because want a lag 
last.ma <- tail(ma.trailing, 1)
ma.trailing.pred <-ts(rep(last.ma, nValid), start = c(1991, nTrain + 1), end = c(1991, nTrain + nValid), freq = 12)
#ma.trailing.pred
plot(train.ts, ylim = c(1300, 2600), ylab = "Ridership", xlab = "Time", bty = "l", xaxt = "n", xlim = c(1991, 2006.25), main = "")
axis(1, at = seq(1991, 2006.25, 1), labels = format(seq(1991, 2006.25, 1)))
#trailing forecast line for the training period
lines(ma.trailing, lwd = 2, col = 4)
#forecast of validation period
lines(ma.trailing.pred, lwd = 2, col = 4, lty = 2)
lines(valid.ts)
#
#legend(1994, 2500, c("Trailing Moving Average", "Trailing MA Forecast", "Validation"), lty = c(1,1,2), lwd = c(1,2,0), col = c(1,4,4), bty = "o")
```

3. Differencing

*$$lag_1$$ is useful to remove the trend. 

*page 85

*There's a lag, $$lag_1$$ is a 1 time lag, the difference between two consecutive values in a series. Doesn't have to be a  $$lag_1$$, can be a $$lag_k$$. A $$lag_7$$ means subtracting from each value
the value on the same day in the previous week. $$lag_1$$ results in a series that measures the changes from one period to the next. 

*For quadratic and exponential trends, often have to apply another round of $$lag_1$$ diffrencing to remove the trend. 

*To remove monthly, seasonality trend, use a $$lag-12$$

*To remove seasonality then trend, $$lag-12$$ then $$lag-1$$


**Seasonality and trend differencing in the Amtrak data.**

```{r}
diff <- diff(diff(ridership.ts, lag = 12), lag = 1)
```


4. Simple Exponential Smoothing

*Like moving average, only for **stationary** data series. 

*Like moving average, but weighted average of all past values so weights decrease exponentially into the past to give more weight to recent data, but not ignore the past altogether. 

*Contains level and error

*Algotithm is learning new level from the latest data. 

*When $\alpha = 1$, no weight is given to early values and the algorithm is not learning anything. Called under smoothing.  

*When $\alpha = 0$ the weight is given exclusively to past values and not to the most recent one. Called over smoothing. 

*In code, if want the function to find the best alpha, **don't suggest one for it**.  

*Like the moving average, simple exponential smoothing should only be used for forecasting series that have **no trend or seasonality**.

*Again, can use differencing to remove seasonality and/or trend and then apply exponential smoothing to the residuals. 

*Below, code to create a simple exponential smoothing forecast where $alpha = 0.2$ applied to twice differenced data. 

```{r}
diff.twice.ts <- diff(diff(ridership.ts, lag = 12), lag = 1)
#validation period
nValid <- 36
#training period of twice differenced data
nTrain <- length(diff.twice.ts) - nValid
#training window of twice differenced data. Start date is modified. 
train.ts <- window(diff.twice.ts, start = c(1992,  2), end = c(1992, nTrain + 1))
#validation window
valid.ts <- window(diff.twice.ts, start = c(1992, nTrain + 2), end = c(1992, nTrain + 1 + nValid))
#simple exponential smoothing model "ANN" on training data
ses <- ets(train.ts, model = "ANN", alpha = .2)
#forecast this model for the length of the validation period
ses.pred <- forecast(ses, h = nValid, level = 0)
ses.pred
#plot the forecast
plot(ses.pred, ylim = c(-250, 300), ylab = "Ridership (twice differenced)", xlab = "Time", bty = "l", xaxt = "n", xlim = c(1991, 2006.25), main = "", flty = 2)
#label the axis
axis(1, at = seq(1991, 2006, 1), labels = format(seq(1991, 2006, 1)))
lines(ses.pred$fitted, lwd = 2, col = "blue")
lines(valid.ts)
```

**Comparison of 2 simple models**

```{r}
ses.opt <- ets(train.ts, model = "ANN")
#forecast this model
ses.opt.pred <- forecast(ses.opt, h = nValid, level = 0)
#accuracy of ses, ANN forecast with alpha = .2
accuracy(ses.pred, valid.ts)
#accuracy of ses, ANN, no predetermined alpha
accuracy(ses.opt.pred, valid.ts)
ses.opt
#parameters function for the alpha and the initial state
ses$par
```

**This is just an aside on how to use dygraphs within a time series.**
```{r}
library(dygraphs)
dygraph(ridership.ts)
forecast <- cbind(ridership.ts, ma.trailing, ma.centered)
dygraph(forecast) %>% dyShading(from="2003-1-1", to="2004-3-1", color="#CCEBD6") %>% dyRangeSelector()

```


5. Advanced Exponential Smoothing

*Can deal with data that has trend and seasonality

* **Additive trend:** the level changes from one period to the next by a fixed amount. 

* **Multiplicative trend:** the level changes from one period to the next by a factor. 

* **Holt's Exponential Smoothing** Contains trend but no seasonality. 

* **Holt Winter's Exponential Smoothing:** Useful for series that contain trend and seasonality. This method is adaptive and allows the level, trend, and seasonality to change over time. 


###This is the AirTravel Data

```{r}
Air <- read.csv("AirTravel.csv")
head(Air)
tail(Air)
range(Air$Air)
air.ts <- ts(Air$Air/10000, start = c(1990, 1), end = c(2000, 12), frequency = 12)
plot(air.ts, ylim = c(2800, 7000), ylab = "Passangers", xlab = "Time", main = "")
```

```{r}
dygraph(air.ts)
nValid <-  24
nTrain <- length(air.ts) - nValid
#moving average by 1 year
ma.centered <- ma(air.ts, order = 12)
#trailing moving average
ma.trailing <- rollmean(air.ts, k = 12, align = "right")
```

