---
title: "Smoothing Methods"
author: "Christine Iyer"
date: "March 9, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

```{r}
library(forecast)
library(zoo)
library(knitr)
ridership <- read.csv("Amtrak.csv")
ridership.ts <- ts(ridership$Ridership_in_thousands, start = c(1991,1), end = c(2004, 3), frequency = 12)
tail(ridership)
```

###2 Types of Moving Averages
**The difference between the two, is the placement of the averaging window over the time series.** 

1. Centered Moving Average

*Averaging operation can suppress seasonality and noise, making the trend more visible.

*If window width is 5, the moving average at t = 3 means averaging time points 1,2,3,4,5. At time point t = 4, the moving average is the average of 2,3,4,5,6. 

*Choosing the window length should be the length of the seasonal cycle. W = 12 for Amtrak data.

*Computed by averaging across data in the past and future of a given time point, therefore not useful for forecasting. 

**Centered moving average uses ma() in forecast package.**

```{r}
# moving average by one year
ma.centered <- ma(ridership.ts, order = 12) 
```


2. Trailing Moving Average

*Useful for forecasting

*Window of width W is placed over the most recent W values of the time series. The k-step ahead forecast is then the average of these W values. 

*The lag makes this method not work for seasonal and trend data. According to textbook, "Seasons with high ridership are under-forecasted, and seasons with low ridership are over-forecasted. A similar issue arises when forecasting a series with a trend: the moving average "lags behind", thereby under-forecasting in the presence of an increasing trend and over-forecasting in the presence of a decreasing trend." But de-trending and de-seasonalizing can be done with regression models, advanced exponential smoothing methods, and differencing. Aftr detrended, etc, forecast, then add back. 

**To do trailing moving aveage, use rollmean() in zoo package**

```{r}
#trailing moving average of entire time series
ma.trailing <- rollmean(ridership.ts, k = 12, align = "right")

```

```{r}
#plot the whole time series
plot(ridership.ts, ylim = c(1300, 2200), ylab = "Ridership", xlab = "Time", bty = "l", xaxt = "n", xlim = c(1991,2004.25), main ="")
#label x axis
axis(1, at = seq(1991, 2004.25, 1), labels = format(seq(1991, 2004.25, 1))) 
#add moving average line
lines(ma.centered, lwd = 2) 
#add trailing moving average line
lines(ma.trailing, lwd = 2, lty = 2) 
#legend for all three lines
legend(1994,2200, c("Ridership","Centered Moving Average", "Trailing Moving Average"), lty=c(1,1,2), lwd=c(1,2,2), bty = "n")

```

**Trailing Moving Average Forecaster**

```{r}
#3 year validation
nValid <- 36
#training period
nTrain <- length(ridership.ts) - nValid
#training window
train.ts <- window(ridership.ts, start = c(1991, 1), end = c(1991, nTrain))
#validation window
valid.ts <- window(ridership.ts, start = c(1991, nTrain + 1), end = c(1991, nTrain + nValid))
#trailing moving average of training period
ma.trailing <- rollmean(train.ts, k = 12, align = "right")
#last value because want a lag 
last.ma <- tail(ma.trailing, 1)

ma.trailing.pred <-ts(rep(last.ma, nValid), start = c(1991, nTrain + 1), end = c(1991, nTrain + nValid), freq = 12)



plot(train.ts, ylim = c(1300, 2600), ylab = "Ridership", xlab = "Time", bty = "l", xaxt = "n", xlim = c(1991, 2006.25), main = "")

axis(1, at = seq(1991, 2006.25, 1), labels = format(seq(1991, 2006.25, 1)))
lines(ma.trailing, lwd = 2, col = "blue")


lines(ma.trailing.pred, lwd = 2, col = "blue", lty = 2)

lines(valid.ts)
```

3. Differencing

*$$lag_1$$ is useful to remove the trend. 

*page 85

*There's a lag, $$lag_1$$ is a 1 time lag, the difference between two consecutive values in a series. Doesn't have to be a  $$lag_1$$, can be a $$lag_k$$. A $$lag_7$$ means subtracting from each value
the value on the same day in the previous week. $$lag_1$$ results in a series that measures the changes from one period to the next. 

*For quadratic and exponential trends, often have to apply another round of $$lag_1$$ diffrencing to remove the trend. 

*To remove monthly, $$lag-12$$

*To remove seasonality then trend, $$lag-12$$ then $$lag-1$$



```{r}
diff <- diff(diff(ridership.ts, lag = 12), lag = 1)
```


4. Simple Exponential Smoothing


*Like moving average, but weighted average of all past values so weights decrease exponentially into the past to give more weight to recent data, but not ignore the past altogether. 

*Like the moving average, simple exponential smoothing should only be used for forecasting series that have no trend or seasonality.

*Again, can use differencing to remove seasonality and/or trend and then apply exponential smoothing to the residuals. 

5. Advanced Exponential Smoothing

*Can deal with data that has trend and seasonality

* **Additive trend:** the level changes from one period to the next by a fixed amount. 

* **Multiplicative trend:** the level changes from one period to the next by a factor. 

* **Hole Winter's Exponential Smoothing:** Useful for series that contain trend and seasonality. This method is adaptive and allows the level, trend, and seasonality to change over time.  


