---
title: "Unit 4 Demo"
author: "Christine Iyer"
date: "March 22, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(forecast)
library(zoo)
```


```{r}

#Set working directory to the file, after reversing the slashes
setwd("C:/Users/Chris Iyer/Documents/AirTravel/data")
#load AirTravel
air <- read.csv("CopyOfAirTravel.csv")
#look at data
head(air)
min(air$Air)
max(air$Air)
head(air)
tail(air)
#create ts object
air.ts <- ts(air$Air, start = c(1990, 1), frequency = 12)
#another way to find min and max values in the ts. Look over to environment to find data
yrange <- range(air.ts)
#plot 
plot(c(1990, 2001), yrange, type = "n", xlab = "Year", ylab = "Air Revenue Passenger Miles (in Millions)", bty = "l", xaxt = "n", yaxt = "n")
#add lines
lines(air.ts, bty = "l")
#x axis at(), seq()
axis(1, at = seq(1990, 2000,1), labels = format(seq(1990, 2000, 1)))
#y axis to format without calculations. las = 2 function is so label is perpendicular to axis
axis(2, at = seq(29000000, 68000000, 5000000), labels = format(seq(29.5, 68, 5.0)), las = 2)
```

**Looking at the plot, we see there is trend and seasonality. Therefore moving average is not the best forecasting method. However centered moving average can be used to explore trend.**

In order to look for trend, will do a centered moving average. 

Step 1. Partition the data

```{r}
validLength <- 12
trainLength <- length(air.ts) - validLength
#create windows
airTrain <- window(air.ts, end = c(1990, trainLength))
airValid <- window(air.ts, start = c(1990, trainLength + 1))
```

Aside, compute a trailing moving average. This is used for forecasting. So it's not appropriate with this dataset, it's just for practice. Use a window size of 12. rollmean()

```{r}
trailingMA <- rollmean(airTrain, k = 12, align = "right")
trailingMA[1]

```

Centered Moving Average. Used for visualization. ma() and order = 12

```{r}
centeredMA <- ma(airTrain, order = 12)
```


**Begin building a plot**
```{r}
#range of y axis
yrange <- range(air.ts)
#basic plot xaxt() and yaxt() suppress the axis
plot(c(1990, 2001), yrange, type = "n", xlab = "Year", ylab = "Air Revenue Passenger Miles (in Millions)", bty = "l", xaxt = "n", yaxt = "n")

#add ts lines, bty = "l" seems to be unnecessary here since used in the previous command. bty = dictates the shape of the box surrounding the plot
lines(air.ts, bty = "l")
#add x axis, remember will forecast out one year so need to extend the axis, las = 2 makes sure all years are labeled, las = 0 skips 1
axis(1, at = seq(1990, 2001, 1), labels = format(seq(1990, 2001, 1)
                                                 ))
#add y axis, reformat numbers.
axis(2, at = seq(29000000, 68000000, 5000000), labels = format(seq(29.5, 68, 5.0)), las = 2)
#add trailing lines
lines(trailingMA, col = "blue")
lines(centeredMA, col = "green")
#add legend, no box around it
legend(1990, 65000000, c("Air Miles Travelled", "Trailing Moving Average", "Centered Moving Average"), lty = c(1,1,1), col = c("black", "blue", "green"), bty = "n")
```

**Centered MA shows a global trend, upward and linear and centered MA is almost always above the forecasting MA. Moving average can't capture seasonality.**


type = "l", "p"
```{r}
plot(c(1990, 2001), c(yrange[1], yrange[2] + 5000000), type = "n", xlab = "Year", ylab = "ARPM in millions", bty = "l", xaxt = "n", yaxt = "n" )

#add training set
lines(air.ts, bty = "l")
#add x axis
axis(1, at = seq(1990, 2001, 1),labels = format(seq(1990, 2001, 1)) )
axis(2, at = seq(29500000, 68000000, 5000000), labels = format(seq(29.5, 68, 5.0)), las = 2)
#most recent forecasted value
maForecastedValue <- tail(trailingMA, 1)
#make this value a ts object
maForecasts <- ts(rep(maForecastedValue, validLength), start=c(1990,trainLength+1), freq=12)
lines(trailingMA, col = "red")
lines(maForecasts, col = "red", lwd = 2, lty = 2)
abline(v = 2000)
arrows(1990, 68500000, 2000, 68500000, code = 3, length = 0.1)
abline(v = 2001)
text(1995, 70000000, "Training")
arrows(2000, 68500000, 2001, 68500000, code = 3, length = 0.1)
text(2000.5, 73000000, "Valid")
```

```{r}
sesAir <- ets(airTrain, model = "ANN", alpha = 0.2)
sesAir
```

```{r}
#set up plot
yrange = range(air.ts)
#plot
plot(c(1990, 2001), c(yrange[1], yrange[2] + 5000000), type = "n", xlab = "ARPM in millions", bty = "l", xaxt = "n", yaxt = "n")
#add ts training set
lines(air.ts, bty = "l")
#add the x axis
axis(1, at = seq(1990, 2001, 1), labels = format(seq(1990, 2001, 1)))
#add y axis
axis(2, at = seq(29500000,68000000,5000000), labels = format(seq(29.5, 68, 5.0)), las = 2)
#add fitted values
lines(sesAir$fitted, col = "red", lwd = 2)
#use forecast function to generate forecasts for the validation period
sesAirPredictions <- forecast(sesAir, h = validLength, level = 0)
#print forecast
sesAirPredictions
#add line to the plot
lines(sesAirPredictions$mean, col = "red", lwd = 2, lty = 2)
#add breakpoints
abline(v = 2000)
arrows(1990, 68500000, 2000, 68500000, code = 3,  length = 0.1)
text(1995, 69500000, "Training")
abline(v= 2001)
arrows(2000, 68500000, 2001, 68500000, code = 3, length = 0.1)
text(2000.5, 69500000, "Valid")
```

**Change the smoothing constant**

```{r}
#set alpha value to 0.8
sesAirBigAlpha <- ets(airTrain, model = "ANN", alpha = 0.8)
sesAirBigAlpha
#plot
yrange = range(air.ts)
#Set up plot
plot(c(1990, 2001), c(yrange[1], yrange[2] + 5000000), type = "n", xlab = "Year", ylab = "ARPM (in millions)", bty = "l", xaxt = "n", yaxt = "n")
#add training set
lines(air.ts, bty = "l")
#add x axis
axis(1, at = seq(1990, 2001, 1), labels = format(seq(1990, 2001, 1)))
axis(2, at = seq(29500000,68000000,5000000), labels = format(seq(29.5, 68, 5.0)), las = 2)
#add fitted values
lines(sesAirBigAlpha$fitted, col = "red", lwd = 2)
#generate a forecast for the validation period
sesAirPredictionsBigAlpha <- forecast(sesAirBigAlpha, h = validLength, level = 0)
lines(sesAirPredictionsBigAlpha$mean, col = "red", lwd = 2, lty = 2)
#add visual breakpoints
abline(v = 2000)
arrows(1990, 68500000, 2000, 68500000, code = 3, length = 0.1)
text(1995, 70000000, "Training")
abline(v = 2001)
arrows(2000, 68500000, 2001, 68500000, code = 3, length = 0.1)
text(2000.5, 70000000, "Validation")

```

**An alpha close to 1, is like the moving average**
Don't set alpha

```{r}
sesAirOpt <- ets(airTrain, model = "ANN")
sesAirOpt
#plot
plot(c(1990, 2001), c(yrange[1], yrange[2] + 5000000), type = "n", xlab = "Year", ylab = "ARMP in millions", bty = "l", xaxt = "n", yaxt = "n")
#add the training set
lines(air.ts, bty = "l")
# Add the x-axis
axis(1, at=seq(1990,2001,1), labels=format(seq(1990,2001,1)))

# Add the y-axis
axis(2, at=seq(29500000,68000000,5000000), labels=format(seq(29.5,68,5.0)), las=2)

#add fitted values to the training set
lines(sesAirOpt$fitted, col = "red", lwd = 2)
#generate forecast of the Opt Alpha
sesAirPredictionsOpt <- forecast(sesAirOpt, h = validLength, level = 0)
#plot the forecast
lines(sesAirPredictionsOpt$mean, col = "red", lwd = 2, lty = 2)
# Add the "visual breakpoints"
abline(v=2000)
arrows(1990, 68500000, 2000, 68500000, code=3, length=0.1)
text(1995, 70000000, "Training")
abline(v=2001)
arrows(2000, 68500000, 2001, 68500000, code=3, length=0.1)
text(2000.5, 73000000, "Validation")
```

###Differencing

```{r}
#detrend air.ts
lag1 <- diff(airTrain, lag = 1)
#deseasonalize
lag12 <- diff(airTrain, lag = 12)
#double differencing
lag1and12 <- diff(diff(airTrain, lag = 1), lag = 12)
#doesn't matter which comes first
# lag12and1 <- diff(diff(airTrain, lag = 12), lag = 1)
# head(lag1and12)
# head(lag12and1)
```

```{r}
par(mfrow = c(2,2))
plot(air.ts)
plot(lag1)
plot(lag12)
plot(lag1and12)
```

###Calculating the point forecast 

```{r}
#mean forecast
meanForecast <- meanf(diff(diff(airTrain, lag = 1), lag = 12), h = 12)
#Moving Average
maForecast <- rollmean(diff(diff(airTrain, lag = 1), lag = 12), k = 12, aligh = "right")
#forecast with the double differenced data
MAForecastforecast <- forecast(maForecast, h = validLength)
#SES with double differenced data
#SESAir <- ets(airTrain, model = "ANN")
SESAir <- ets(diff(diff(airTrain, lag = 1), lag = 12), model = "ANN")
MeanForecastforecast <- forecast(meanForecast, h = validLength)
SESAirForecast <- forecast(SESAir, h = validLength, level = 0)
MeanForecastforecast
MAForecastforecast
SESAirForecast$mean
```

**Reseaonalize the meanf()**

```{r}
realMeanForecasts <- vector()
for (i in 1:validLength){
  if(i == 1){
    
    realMeanForecasts[i] <- meanForecast$mean [i]  + airTrain[(trainLength+i)-validLength] + (airTrain[trainLength] - airTrain[trainLength - validLength])
  } else{
    realMeanForecasts[i] <- meanForecast$mean[i] + airTrain[(trainLength+i)-validLength] + (realMeanForecasts[i-1] - airTrain[trainLength+i-1-validLength])
  }
}
#actual MA forecast
realMeanForecasts


```


```{r}
#reseasonalize and trend
realForecasts <- vector()
for (i in 1:validLength){
  if(i == 1){
    
    realForecasts[i] <- MAForecastforecast$mean[i]  + airTrain[(trainLength+i)-validLength] + (airTrain[trainLength] - airTrain[trainLength - validLength])
  } else{
    realForecasts[i] <- MAForecastforecast$mean[i] + airTrain[(trainLength+i)-validLength] + (realForecasts[i-1] - airTrain[trainLength+i-1-validLength])
  }
}
#actual MA forecast
realForecasts
```

```{r}
SESrealForecasts <- vector()
for (i in 1:validLength){
  if(i == 1){
    
    SESrealForecasts[i] <- SESAirForecast$mean [i]  + airTrain[(trainLength+i)-validLength] + (airTrain[trainLength] - airTrain[trainLength - validLength])
  } else{
    SESrealForecasts[i] <- SESAirForecast$mean[i] + airTrain[(trainLength+i)-validLength] + (SESrealForecasts[i-1] - airTrain[trainLength+i-1-validLength])
  }
}
#actual SES forecast
SESrealForecasts
```

**Plot the MA forecast against the Actual**

```{r}
yrange <- range(air.ts)
plot(c(1990, 2001), c(yrange[1], yrange[2]+5000000), type="n", xlab="Year", ylab="Air Revenue Passenger Miles (millions)", bty="l", xaxt="n", yaxt="n")
lines(air.ts, bty = "l")
# Add the x-axis
axis(1, at=seq(1990,2001,1), labels=format(seq(1990,2001,1)))

# Add the y-axis
axis(2, at=seq(29500000,68000000,5000000), labels=format(seq(29.5,68,5.0)), las=2)
lines(seq(2000, 2001-1/validLength, 1/validLength), realForecasts, col="red", lwd=2, lty=2)

lines(seq(2000, 2001-1/validLength, 1/validLength), SESrealForecasts, col="blue", lwd=2, lty=2)
# Add the "visual breakpoints"
abline(v=2000)
arrows(1990, 68500000, 2000, 68500000, code=3, length=0.1)
text(1995, 70000000, "Training")
abline(v=2001)
arrows(2000, 68500000, 2001, 68500000, code=3, length=0.1)
text(2000.5, 73000000, "Validation")


```

**Re-Seasonalized and Re-Trended Forecasts**

```{r}
Actual <- tail(air$Air, 12)
Actual
yrange <- range(Actual)

plot(c(1, 12), c(yrange[1], yrange[12]+5000000), type="n", xlab="Month", ylab="Air Revenue Passenger Miles (millions)", bty="l", xaxt="n", yaxt="n")
axis(1, at=seq(1, 12), labels=format(seq(1, 12, 1), ylab = "Month"))

# Add the y-axis
axis(2, at=seq(30000000,98000000,5000000), labels=format(seq(30,98,5.0)), las=2)
lines(Actual, type="l", bty="l", lwd = 2)
lines(realForecasts, type="l", bty="l", col = "blue", lwd = 2)
lines(SESrealForecasts, type="l", bty="l", col = "red", lwd = 2)
lines(realMeanForecasts, type="l", bty="l", col = "orange", lwd = 2)
legend(10, 45000000, c("Actual", "MA", "Meanf()", "SES"), lty = c(1,1,1,1), col = c("black", "blue", "orange", "red"), bty = "n", lwd = c(2,2,2,2))

```

```{r}
range(air.ts)
```


