---
title: "Week13"
author: "Christine Iyer"
date: "April 19, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For predicting whether the agricultural epidemic of powdery mildew in mango will erupt in a certain year in the state of Uttar Pradesh in India, Misra et al. (2004) used annual outbreak 9 A. K. Misra, O. M. Prakash, and V. Ramasubramanian. Forewarning powdery mildew caused by Oidium mangiferae in mango (Mangifera indica) using logistic regression models. Indian Journal of Agricultural Science, 74(2):84-87, 2004 records during 1987-2000. The epidemic typically occurs in the third and fourth week of March, and hence outbreak status is known by the end of March of a given year. The authors used a logistic regression model with two weather predictors (maximum temperature and relative humidity) to forecast an outbreak. The data is shown in the table below and are available in PowderyMildewEpidemic.xls.

```{r}
library(pander)
library(forecast)
library(caret)
library(dplyr)
mildew <- read.csv("PowderyMildewEpidemic.csv")
mildew <- mildew %>% mutate(Outbreak1 = ifelse(Outbreak == "Yes", 1,0))
pander(head(mildew))
plot(mildew$MaxTemp ~ mildew$RelHumidity, xlab = "Relative Humidity", ylab = "Max Temp (degrees Celsius)", col = mildew$Outbreak1 + 1, bty = "l", pch = 15)
legend(60,29, c("No Outbreak", "Outbreak"), col = 1:2, pch = 15, bty = "l")
```

```{r}
plot(mildew$RelHumidity ~ mildew$MaxTemp, ylab = "Relative Humidity", xlab = "Max Temp (degrees Celsius)", col = mildew$Outbreak1 + 1, bty = "l", pch = 15)
legend(27,70, c("No Outbreak", "Outbreak"), col = 1:2, pch = 15, bty = "l")
```


1. In order for the model to serve as a forewarning system for farmers, what requirements must be satisfied regarding data availability? 

2. Write an equation for the model fitted by the researchers in the form of equation (8.1). Use predictor names instead of x notation. 

3. Create a scatter plot of the two predictors, using different hue for epidemic and non-epidemic markers. Does there appear to be a relationship between epidemic status and the two predictors?

4. Compute naive forecasts of epidemic status for years 1995- 1997 using next-year forecasts (Ft+1 = Ft). What is the naive forecast for year 2000? Summarize the results for these four years in a classification matrix. Use a roll forward naive forecast.

5. Partition the data into training and validation periods, so that years 1987-1994 are the training period. Fit a logistic regression to the training period using the two predictors, and report the outbreak probability as well as a forecast for year 1995 (use a threshold of 0.5).

