---
title: "Unit 4 Demo"
author: "Christine Iyer"
date: "March 22, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(forecast)
library(zoo)
```


```{r}

#Set working directory to the file, after reversing the slashes
setwd("C:/Users/Chris Iyer/Documents/AirTravel/data")
#load AirTravel
air <- read.csv("CopyOfAirTravel.csv")
#look at data
head(air)
min(air$Air)
max(air$Air)
head(air)
tail(air)
#create ts object
air.ts <- ts(air$Air, start = c(1990, 1), frequency = 12)
#another way to find min and max values in the ts. Look over to environment to find data
yrange <- range(air.ts)
#plot 
plot(c(1990, 2001), yrange, type = "n", xlab = "Year", ylab = "Air Revenue Passenger Miles (in Millions)", bty = "l", xaxt = "n", yaxt = "n")
#add lines
lines(air.ts, bty = "l")
#x axis at(), seq()
axis(1, at = seq(1990, 2000,1), labels = format(seq(1990, 2000, 1)))
#y axis to format without calculations. las = 2 function is so label is perpendicular to axis
axis(2, at = seq(29000000, 68000000, 5000000), labels = format(seq(29.5, 68, 5.0)), las = 2)
```

**Looking at the plot, we see there is trend and seasonality. Therefore moving average is not the best forecasting method. However centered moving average can be used to explore trend.**

In order to look for trend, will do a centered moving average. 

Step 1. Partition the data

```{r}
nValid <- 12
nTrain <- length(air.ts) - nValid
#create windows
airTrain <- window(air.ts, end = c(1990, nTrain))
airValidate <- window(air.ts, start = c(1990, nTrain + 1))
```

Aside, compute a trailing moving average. This is used for forecasting. So it's not appropriate with this dataset, it's just for practice. Use a window size of 12. rollmean()

```{r}
trailingAir <- rollmean(airTrain, k = 12, align = "right")
trailingAir[1]

```

Centered Moving Average. Used for visualization. ma() and order = 12

```{r}
centeredAir <- ma(airTrain, order = 12)
```


**Begin building a plot**
```{r}
#range of y axis
yrange <- range(air.ts)
#basic plot xaxt() and yaxt() suppress the axis
plot(c(1990, 2001), yrange, type = "n", xlab = "Year", ylab = "Air Revenue Passenger Miles (in Millions)", bty = "l", xaxt = "n", yaxt = "n")

#add ts lines, bty = "l" seems to be unnecessary here since used in the previous command. bty = dictates the shape of the box surrounding the plot
lines(air.ts, bty = "l")
#add x axis, remember will forecast out one year so need to extend the axis, las = 2 makes sure all years are labeled, las = 0 skips 1
axis(1, at = seq(1990, 2001, 1), labels = format(seq(1990, 2001, 1)
                                                 ))
#add y axis, reformat numbers.
axis(2, at = seq(29000000, 68000000, 5000000), labels = format(seq(29.5, 68, 5.0)), las = 2)
#add trailing lines
lines(trailingAir, col = "blue")
lines(centeredAir, col = "green")
#add legend, no box around it
legend(1990, 65000000, c("Air Miles Travelled", "Trailing Moving Average", "Centered Moving Average"), lty = c(1,1,1), col = c("black", "blue", "green"), bty = "n")
```

**Centered MA shows a global trend, upward and linear and centered MA is almost always above the forecasting MA. Moving average can't capture seasonality.**