---
title: "Cheat Sheet"
author: "Christine Iyer"
date: "April 13, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[For reference](https://github.com/derekfranks/practice_assignment/blob/master/practice_assignment.rmd)

**Download and unzip a file**

```{r}
dataset_url <- "http://s3.amazonaws.com/practice_assignment/diet_data.zip"
download.file(dataset_url, "diet_data.zip")
unzip("diet_data.zip", exdir = "diet_data")
list.files("diet_data")
#look at one file
andy <- read.csv("diet_data/Andy.csv")
#find his final weight on Day 30 4 ways
andy[30, "Weight"]
andy[which(andy$Day == 30), "Weight"]
andy[which(andy[,"Day"] == 30), "Weight"]
subset(andy$Weight, andy$Day==30)
#net loss
andy[30, "Weight"] - andy[1, "Weight"]
#list all files as an object
files <- list.files("diet_data")
files[3:5]
#In file(file, "rt") :cannot open file 'John.csv': No such file or directory
#head(read.csv(files[3]))
files_full <- list.files("diet_data", full.names = T)
#see head of John
head(read.csv(files_full[3]))
#combine andy and david
andy_david <- rbind(andy, read.csv(files_full[2]))
#subset of day 25
day_25 <- andy_david[which(andy_david$Day ==25), ]
#using a loop to combine everyones, so don't have to manually append each one
#create an object 
dat <- data.frame()
#loop
for (i in 1:5) {
        dat <- rbind(dat, read.csv(files_full[i]))
}
#media weight on day 30
median(dat$Weight, na.rm=TRUE)
#subset day 30
dat_30 <- dat[which(dat[, "Day"] == 30),]
#median of that subset
median(dat_30$Weight)
#the values are different because someone was excluded in the prior function
#build a function to see the median weight on any given day



```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=60)}
weight_median <- function(directory, day){
  files_list <- list.files(directory, full.names = T)
  dat <- data.frame()
  for(i in 1:5){
    dat <- rbind(dat, read.csv(files_list[i]))
  }
 dat_subset <- dat[which(dat[,"Day"]==day),]
 median(dat_subset[,"Weight"], na.rm = T)
}

```

```{r}
weight_median(directory = "diet_data", day = 20)
weight_median("diet_data", 4)
weight_median("diet_data", 17)
```

**for() loop

```{r}
x <- c("a", "b", "c", "d", "e", "f")
for(i in 1:6){
  print(x[i])
}
```

**seq_along() and for() loop**

```{r}
for(i in seq_along(x)){
  print(x[i])
}
```

**using letter notation**

```{r}
for(letter in x){
  print(letter)
}
```

**simplest, no curly braces**

```{r}
for(i in 1:4) print(x[i])
for(i in 1:6) print(x[i])
```

**Nested for loops**
for loop inside a for loop, useful in a 2 dimensional matrix where want to loop over the rows and then columns

```{r}
#create matrix. matrix(data to include, #rows, #columns)
x <- matrix(1:6, 2,3)
x
for(i in seq_len(nrow(x))){
  for(j in seq_len(ncol(x))){
    print(x[i,j]/2)
    print(x[i,j]/4)
  }
}
```

**while loop**
make sure the condition sets an end point

```{r}
z <- 5
while(z >= 3 && z <=10){
  print(z)
  coin <- rbinom(1,1,0.5)
  #no end point yet
  
  if(coin == 1){#random walk
    z <- z + 1
  }
  else{
    z <- z-1
  }
}
```

**repeat()**
initiates an infinite loop, the only way to exit is to call a break

```{r}
x0 <- 1
tol <- 1e-8

repeat{
  x1 <- rbinom(1,1,0.5)
  
  if(abs(x1-x0) < tol){
    break
  } else{
    x0 <- x1
  }
}
```

**skiping an iteration, use next()**

```{r}
for(i in 1:50){
  if(i <= 20){
    next
  }
  #Do something here
}
```

**Writing Functions**
adding 2 numbers

```{r}
#2 values so 2 arguments
add2 <- function(x,y){
  x+y
}
#now test it to return 2+3
add2(4,3)
```

take a vector of numbers and return a subset of it that's above 10

```{r}
above10 <- function(x){
  
  use <- x>10
  x[use]
} 
#x <- c(5,45,34,23)
above10(x)
```

```{r}
above <- function(x, n = 10){
  use <- x > 10
  x[use]
  
}
above(14)
above(4)
```

```{r}
setwd("/Users/Chris Iyer/Documents/")
ClinicAb <- read.csv("Clinics.csv")


columnmean <- function(x){
  nc <- ncol(x)
  means <- numeric(nc)
  for(i in 1:nc){
    means[i] <- mean(x[, i])
  }
  means
}
columnmean(ClinicAb)
```

